Dynamics of innovation in stigmergic collectives

### 2023-06-13 LaTeX/draft.tex
Remark
MoreEdits

Mention that the TD(0)-learning results in
\[
    v(s_c) \leftarrow (1-\alpha) v(s_c) + \alpha r(s^*)
\]
as $v(s^*) = 0$.

@@

Let us consider an agent whose orientation $\ph = ( \cos \theta, \sin \theta)$ wants to relax to a
neutral orientation with stochasticity associated with estimation error. The dynamics of the orientation
can be written as a stochastic differential equation given by
\begin{align}
    \dot{\theta}(t) =& \ - \alpha \theta(t) + \sqrt{2 D} \eta(t), \\
    \langle \eta(t) \eta(0) \rangle =& \ \delta(t).
\end{align}
This is the evolution of the famous Ornstein-Uhlenbeck process which has been extremely well studied
and the evolution of the probability density can be solved exactly. The Fokker-Planck equation for
its dynamics which provides the probability of a given state $\theta(t)$ i.e. $p(\theta, t)$, is given by
\[
    \frac{\partial \P}{\partial t} = \alpha \frac{\partial(\theta \P)}{\partial \theta }
    + D \frac{\partial^2 \P}{\partial \theta^2}
\]
The dynamics of the angle relaxes to equilibrium angle $\theta = 0$ in the absence of fluctuations.
The evolution of the mean is $\dot{h}(t) = - \alpha h(t)$ where $h(t) = \E [ \theta(t) ]$
while the co-variance dynamics is given by $\dot{\zeta}(t) = -2\alpha \zeta(t) + 2D$ with
$\zeta = \E [  {(\theta(t) - h(t))}^2 ] = \E [  \theta^2(t) ] - h{(t)}^2$.

The position of the agent evolves as $\dot{\r}(t) = v_o \ph(t)$. Under the assumption that the angle
is small, we can write $\dot{\r}(t) = v_o (1, \theta)$. Thus we have $\yd (t) = v_o \theta$ which can
be used to rewrite the orientation equation as
\begin{align}
    \ddot{y}(t) =& \ - \alpha \yd (t) + v_o\sqrt{2 D} \eta(t).
\end{align}
This second order equation needs to boundary condition. We require $y(0) = 0$ and $\yd (0) = 0$ which
comes from setting the initial orientation to be 0. We are interested in the evolution of the mean and
the variance of this SDE. Let $m(t)= \E [  y(t) ]$ where the averaging is the ensemble average
(equivalently the average of the several trajectories of the same dynamics). The mean $m(t)$ evolves as
\begin{align}
    \ddot{m}(t) =& \ -\alpha \dot{m}(t).
\end{align}
By applying the above boundary condition, we can immediately see that $m(t) = 0$. Nevertheless, we are
interested in the evolution of the standard deviation, $\beta(t) = \E [  {(y(t) - m(t))}^2 ]$.
In order to derive an expression for that, we differentiate the definition once to get
\begin{align*}
    \dot{\beta}(t) =& \ 2 \E [ \{ y(t) - m(t) \} \{ \yd(t) - \dot{m}(t) \} ], \\
    = & \ 2 \E [ y \yd ] - 2 m \dot{m}, \\
    = & \ 2 v_o \{ \E [ y \theta ] - m \dot{m}\}.
\end{align*}
We can estimate $\E [ y \theta ]$ as follows
\begin{align*}
    \E [ y \theta ] =& \ v_o \E \bigg[ \int_0^t \theta(s) \theta(t) \ \d s \bigg], \\
    =& \ v_o  \int_0^t \E [ \theta(t) \theta(s) ] \ \d s.
\end{align*}
We know that for an OU process, 
\[
    \E [ \theta(t) \theta(s) ] = \frac{D}{\alpha} \bigg[ \e^{-\alpha(t-s)}-\e^{-\alpha(t+s)} \bigg].
\]
From this we see that
\begin{align*}
    \E [ y \theta ] =& \ \frac{v_o D}{\alpha} \int_0^t  \bigg[ \e^{-\alpha(t-s)}-\e^{-\alpha(t+s)} \bigg] \ \d s, \\
    =& \ \frac{v_o D}{\alpha^2} \bigg[ 1 - 2 \e^{-\alpha t} + \e^{-2 \alpha t}\bigg].
\end{align*}
We thus have the evolution of the co-variance of $y(t)$ as
\begin{align}
    \dot{\beta}(t) = & \  \frac{2 v_o^2 D}{\alpha^2} \bigg[ 1 - 2 \e^{-\alpha t} + \e^{-2 \alpha t}\bigg].
\end{align}
We finally have the variance as a function of time as
\begin{align}
    \beta(t) & = \frac{v_o^2 D}{\alpha^3} \bigg[ 2 \alpha t - \e^{-2 \alpha t} + 4 \e^{-\alpha t} - 3 \bigg].
\end{align}
For $t \gg \alpha^{-1}$ we can immediately see that $\beta(t) \approx 2v_o^2 D t/\alpha^2$. We are interested
in the variance at time, $t^* = c/v_o$ where $c$ is the distance of the agent from the goal A and $v_o$
is the intrinsic speed of the agent. At this time the location of the agent along x-axis is $x(t^*)=c$
while the variance in $y(t)$ is $\beta(t^*) \approx 2v_o D c/\alpha^2$.
% Differentiating this once more, we get
% \begin{align*}
%     \ddot{\beta}(t) =& \ 2 [ v_o \langle (\yd \theta + y \dot{\theta} \rangle - \dot{m}^2 - m \ddot{m}]
% \end{align*}

>>>>

Mention that the TD(0)-learning results in
\[
    v(s_c) \leftarrow (1-\alpha) v(s_c) + \alpha r(s^*)
\]
as $v(s^*) = 0$. Mention that we don't need $q(s, a)$ as the value function is parameterized in a fashion to account for
actions.

@@

Let us consider an agent whose orientation $\ph = ( \cos \theta, \sin \theta)$ wants to relax to a
neutral orientation with stochasticity associated with estimation error. The dynamics of the orientation
can be written as a stochastic differential equation given by
\begin{align}
    \dot{\theta}(t) =& \ - \alpha \theta(t) + \sqrt{2 D} \eta(t), \\
    \langle \eta(t) \eta(0) \rangle =& \ \delta(t).
\end{align}
This is the evolution of the famous Ornstein-Uhlenbeck process which has been extremely well studied
and the evolution of the probability density can be solved exactly. The Fokker-Planck equation for
its dynamics which provides the probability of a given state $\theta(t)$ i.e. $p(\theta, t)$, is given by
\[
    \frac{\partial \P}{\partial t} = \alpha \frac{\partial(\theta \P)}{\partial \theta }
    + D \frac{\partial^2 \P}{\partial \theta^2}
\]
The dynamics of the angle relaxes to equilibrium angle $\theta = 0$ in the absence of fluctuations.
The evolution of the mean is $\dot{h}(t) = - \alpha h(t)$ where $h(t) = \E [ \theta(t) ]$
while the co-variance dynamics is given by $\dot{\zeta}(t) = -2\alpha \zeta(t) + 2D$ with
$\zeta = \E [  {(\theta(t) - h(t))}^2 ] = \E [  \theta^2(t) ] - h{(t)}^2$. We can solve this to get
the time-evolution of the co-variance as $\zeta(t) = D/\alpha(1-\e^{-2\alpha t})$.

The position of the agent evolves as $\dot{\r}(t) = v_o \ph(t)$. Under the assumption that the angle
is small, we can write $\dot{\r}(t) = v_o (1, \theta)$. Thus we have $\yd (t) = v_o \theta$ which can
be used to rewrite the orientation equation as
\begin{align}
    \ddot{y}(t) =& \ - \alpha \yd (t) + v_o\sqrt{2 D} \eta(t).
\end{align}
This second order equation needs two boundary condition. We require $y(0) = 0$ and $\yd (0) = 0$ which
comes from setting the initial orientation to be 0. We are interested in the evolution of the mean and
the variance of this SDE. Let $m(t)= \E [  y(t) ]$ where the averaging is the ensemble average
(equivalently the average of the several trajectories of the same dynamics). The mean $m(t)$ evolves as
\begin{align}
    \ddot{m}(t) =& \ -\alpha \dot{m}(t).
\end{align}
By applying the above boundary condition, we can immediately see that $m(t) = 0$. Nevertheless, we are
interested in the evolution of the standard deviation, $\beta(t) = \E [  {(y(t) - m(t))}^2 ]$.
In order to derive an expression for that, we differentiate the definition once to get
\begin{align*}
    \dot{\beta}(t) =& \ 2 \E [ \{ y(t) - m(t) \} \{ \yd(t) - \dot{m}(t) \} ], \\
    = & \ 2 \E [ y \yd ] - 2 m \dot{m}, \\
    = & \ 2 v_o \{ \E [ y \theta ] - m \dot{m}\}.
\end{align*}
We can estimate $\E [ y \theta ]$ as follows
\begin{align*}
    \E [ y \theta ] =& \ v_o \E \bigg[ \int_0^t \theta(s) \theta(t) \ \d s \bigg], \\
    =& \ v_o  \int_0^t \E [ \theta(t) \theta(s) ] \ \d s.
\end{align*}
We know that for an OU process, 
\[
    \E [ \theta(t) \theta(s) ] = \frac{D}{\alpha} \bigg[ \e^{-\alpha(t-s)}-\e^{-\alpha(t+s)} \bigg].
\]
From this we see that
\begin{align*}
    \E [ y \theta ] =& \ \frac{v_o D}{\alpha} \int_0^t  \bigg[ \e^{-\alpha(t-s)}-\e^{-\alpha(t+s)} \bigg] \ \d s, \\
    =& \ \frac{v_o D}{\alpha^2} \bigg[ 1 - 2 \e^{-\alpha t} + \e^{-2 \alpha t}\bigg].
\end{align*}
We thus have the evolution of the co-variance of $y(t)$ as
\begin{align}
    \dot{\beta}(t) = & \  \frac{2 v_o^2 D}{\alpha^2} \bigg[ 1 - 2 \e^{-\alpha t} + \e^{-2 \alpha t}\bigg].
\end{align}
We finally have the variance as a function of time as
\begin{align}
    \beta(t) & = \frac{v_o^2 D}{\alpha^3} \bigg[ 2 \alpha t - \e^{-2 \alpha t} + 4 \e^{-\alpha t} - 3 \bigg].
\end{align}
For $t \gg \alpha^{-1}$ we can immediately see that $\beta(t) \approx 2v_o^2 D t/\alpha^2$. We are interested
in the variance at time, $t^* = c/v_o$ where $c$ is the distance of the agent from the goal A and $v_o$
is the intrinsic speed of the agent. At this time the location of the agent along x-axis is $x(t^*)=c$
while the variance in $y(t)$ is $\beta(t^*) \approx 2v_o D c/\alpha^2$.
% Differentiating this once more, we get
% \begin{align*}
%     \ddot{\beta}(t) =& \ 2 [ v_o \langle (\yd \theta + y \dot{\theta} \rangle - \dot{m}^2 - m \ddot{m}]
% \end{align*}
